---
title: "Pairwise comparisons for each line pairs across each timepoint."
format: html
date: "2026-03-17"
bibliography: refs-stat-anal.bib
execute:
  eval: false
  fig-format: png
  fig.dpi: 300
---

Built with R version `r getRversion()`

## Rationale

Using only subset of data for SAM and design matrix to extract interesting comparisons.

We have four days, two of which are cold (c1, c2) and two of which are warm (r1, r2 "regrowth") with one-day break (b).

c1 - c2 - b - r1 - r2

:::{.callout-note}
For *historical* reasons the following names will be used

d1 - d2 - b - d3 - d4
:::

In each day we have three samples: dawn, day, and dusk.

c1[dawn, day, dusk] - c2[dawn, day, dusk] - b - r1[dawn, day, dusk] - r2[dawn, day, dusk]

And this will be recoded as 1 - 12.
So in all there are 12 time points. This scheme is repeated for four maize inbred lines.

To check if there are time-dependent changes between lines we test the timepoint:line interaction with factorial model as described in [edgeR, p. 41](https://bioconductor.org/packages//release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf).


`DESeq2` package will be used [@loveModeratedEstimationFold2014].

## Environment cleaning and set-up

```{r setup, include = FALSE}
# load required packages
library(DESeq2)
library(GenomicRanges)
library(BiocParallel)
```

```{r install packages}
#| eval: false
#| output: false
#| message: false
# Install required packages if not done previously
## From Bioconductor
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("GenomicRanges")
BiocManager::install("DESeq2")
BiocManager::install("genefilter")
BiocManager::install("apeglm")

## From CRAN
install.packages("pheatmap")
install.packages("RColorBrewer")
install.packages("viridis")
install.packages("PoiClaClu")
install.packages("glmpca")
install.packages("ggplot2")
install.packages("dplyr")
install.packages("ashr")
```

```{r}
#| eval: false
#| output: false

# Clean working directory
rm(list=ls())
```

## Data preprocessing

### Load raw count data

Full dataset for cold + regrowth

```{r read data, cache=TRUE}
load("../../../input-data/rdata/counts_and_metadata.RDa")

# previously prepared range data
load("../../../input-data/rdata/ranges.RDa")

# Check if files are in concert
all(rownames(coldata) %in% colnames(cts))
all(rownames(coldata) == colnames(cts))
```

### Subset SAM data

Subsetting is done at the level of raw data (counts and coldata).
```{r subset SAM data}
# subset coldata to contain only sam data
coldata.sam=subset(coldata, tissue=="sam")

# check
dim(coldata.sam)

# add variable describing chronology of samples
coldata.sam$timepoint = as.integer(as.numeric(coldata.sam$time)) + ( as.integer(as.numeric(coldata.sam$day)) - 1 ) * 3
# make 'timepoint' a factor
coldata.sam$timepoint = as.factor(coldata.sam$timepoint)

# Adding grouping variable to easily analyse all comparisons needed
coldata.sam$group <- factor(paste0(coldata.sam$line, "_", coldata.sam$timepoint))

# vector of sample names to keep - needed to select columns from count file
keep.sam=rownames(coldata.sam)

# check - only samples named "s..." should remain
keep.sam
length(keep.sam)

# subsetting counts
cts.sam=subset(cts, select=keep.sam)

# check
dim(cts.sam)
colnames(cts.sam)

# Check if counts and coldata are in concert
all(rownames(coldata.sam) %in% colnames(cts.sam))
all(rownames(coldata.sam) == colnames(cts.sam))

keep = c("coldata.sam", "cts.sam", "gzakresy")

rm(list = setdiff(ls(), keep))
```

### Create DESeq2 object for SAM only

```{r create DESeq2 object, cache=TRUE}
library("DESeq2")
#| output: false
#| message: false

# Design according to:
# https://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#time-course-experiments

dds=DESeqDataSetFromMatrix(countData=cts.sam, colData=coldata.sam, rowRanges=gzakresy, design= ~ 0 + group)

```

### Pre-filtering

```{r pre-filtering, cache=TRUE}
# at least 3 samples with a count of 10 or higher
keep <- rowSums(counts(dds) >= 10) >= 3
dds.f <- dds[keep,]

# check
dim(dds.f)

# what percent of genes remained?
100*(dim(dds.f)[1]/dim(dds)[1])

keep = c("dds.f")

rm(list = setdiff(ls(), keep))
```

## Analysis

### Test


```{r analysis, cache=TRUE}
#| eval: false

# parallelization setup
## otherwise test runs very long

register(MulticoreParam(8))

# analysis using simple design already specified in dds.f
dds.f <- DESeq(dds.f, parallel = TRUE)
```

```{r env cleanup}
# genes from the following object will be used as background in GO enrichment analysis
save(dds.f, file="../rdata-saved/dds.f.RDa")
```

### Selecting interesting comparisons

```{r refresh}
# remove all objects and load only dds.f object (result of DEseq function)
rm(list = ls())

load("../rdata-saved/dds.f.RDa")
```

For logFC shrinkage `ashr` is used [@stephensFalseDiscoveryRates2016] instead of `apeglm` as the latter doesn't support contrasts.

```{r}
# according to post by ATpoint
# https://support.bioconductor.org/p/9139197/#9139238

unique_combn <- combn(c("a554", "s018", "s84", "s311"), 2)

timepoints = c(1 : 12)

allcontrasts = list()

for (timepoint in timepoints){
  contrasts <- lapply(1:ncol(unique_combn), function(x){
  
    c(paste0("group", unique_combn[1,x], "_", timepoint), paste0("group", unique_combn[2,x], "_", timepoint))
  
  })
  
  allcontrasts = append(allcontrasts, contrasts)

}

invisible(
  lapply(allcontrasts, function(contr) {

    contr1 = unlist(contr)
    
    
    res_name <- paste(gsub("group", "", contr1[1]), gsub("group", "", contr1[2]), sep=".")
    
    res=lfcShrink(dds.f, contrast=list(contr1[1], contr1[2]), type="ashr")
    
    assign(res_name, res, envir = .GlobalEnv)
  })
)
```

```{r save-contr}
to_save = ls(pattern = "\\.[sa]")

save(list = to_save, file="../rdata-saved/contr_pair_sam.RDa")

rm(list = setdiff(ls(), to_save))
```

```{r sigs}
# extract and save only significant genes
original_names = ls(pattern = "\\.[sa]")

invisible(
  lapply(original_names, function(name) {
    obj <- get(name)

    sig_name = paste0(name, "_sam_sig")
    
    res <- obj[which(obj$padj < 0.05), ]
    assign(sig_name, res, envir = .GlobalEnv)

  })
)

to_save = ls(pattern = "sig$")

# this file will be used for clustering with mfuzz
save(list = to_save, file="../rdata-saved/contr_pair_sam_sig005.RDa")

#rm(list = setdiff(ls(), to_save))
```

## References
