---
title: "Join tables summarising significant genes in clusters, plus annotation"
format: html
date: "2026-03-13"
bibliography: ../../../refs-stat-anal.bib
execute:
  fig-format: png
  fig.dpi: 300
---

This code used files made in:
- [13-SAM-genes-cluster-table.qmd](13-SAM-genes-cluster-table.qmd) - results for SAM in cold
- [13-leaf-genes-cluster-table.qmd](../../leaf/scripts/13-leaf-genes-cluster-table.qmd) - results for leaf in cold
- [13-SAM-ctrl-genes-cluster-table.qmd](../../../../project1-ctrl-stat/vs-day-one/sam/scripts/13-SAM-ctrl-genes-cluster-table.qmd) - results for SAM in control
- [13-leaf-ctrl-genes-cluster-table.qmd](../../../../project1-ctrl-stat/vs-day-one/leaf/scripts/13-leaf-ctrl-genes-cluster-table.qmd) - results for leaf in control

Built with R version `r getRversion()`

## R Preliminaries

Environment cleaning and set-up

```{r}
#| output: false

# Clean working directory
rm(list=ls())
```

Load `DESeq2` results - tables for contrasts, saved in [02-SAM-only-loop-regr-cold.qmd](02-SAM-only-loop-regr-cold.qmd)

```{r load}
library(DESeq2)

load("../rdata-saved/sam_cold.DE_cluster_table.RDa")
load("../../leaf/rdata-saved/leaf_cold.DE_cluster_table.RDa")
load("../../../../project1-ctrl-stat/vs-day-one/sam/rdata-saved/sam_ctrl.DE_cluster_table.RDa")
load("../../../../project1-ctrl-stat/vs-day-one/leaf/rdata-saved/leaf_ctrl.DE_cluster_table.RDa")
```

Join together log2FC values for significant genes in each comparison.

```{r join}
# !! "unquoting" operator enforces immediate evaluation of the left-side variable inside the function.
# := "glue" operator (frequently used instead of = in rename), allows dynamical mapping of columns' names.

library(dplyr)
library(tibble)
in_names <- c("leaf_cold", "leaf_ctrl", "sam_cold", "sam_ctrl")

fc_decl <- data.frame(Name = character())

for (name in in_names) {
  in_obj=as.data.frame(get(name))

  fc_decl = full_join(fc_decl, in_obj, by = "Name")
}

#rm(list = in_names)
```

Add annotation.

```{r anno}
# load annotation
anno = read.table("../../../input-data/go-uniprot/NAM_descr_best", header=FALSE, stringsAsFactors = FALSE, sep="\t", quote = "")
colnames(anno) <- c("Name", "description")

# join with DE-data
fc_decl_anno <- left_join(fc_decl, anno, by = "Name")

# join with UPs
ups = read.table("../../../input-data/go-uniprot/NAM2UP_all_forGOextraction", header=FALSE, stringsAsFactors = FALSE, sep="\t", quote = "")
colnames(ups) <- c("Name", "UniProt")

fc_decl_anno_up <- left_join(fc_decl_anno, ups, by = "Name") |>
  relocate(UniProt, .after = Name)
```

Save to file

```{r save}
write.table(fc_decl_anno_up, file = "../../genes_cluster_table_coldctrl.csv", quote = FALSE, sep = "\t", row.names = FALSE)
```