---
title: "Table summarising significant genes in clusters"
format: html
date: "2026-03-13"
bibliography: ../../../refs-stat-anal.bib
execute:
  fig-format: png
  fig.dpi: 300
---

The aim of this code is to make table with log2FC values for significant genes in each comparison.
This data will be joined with cluster number.

This will be done for each project-tissue combination and eventually all four tables will be joined together.
In the end gene annotation will be added

Built with R version `r getRversion()`

## R Preliminaries

Environment cleaning and set-up

```{r}
#| output: false

# Clean working directory
rm(list=ls())
```

Load `DESeq2` results - tables for contrasts, saved in [02-SAM-only-loop-regr-cold.qmd](02-SAM-only-loop-regr-cold.qmd)

```{r load}
library(DESeq2)

load("../rdata-saved/sigs_all_SAM1.RDa")
```

Join together log2FC values for significant genes in each comparison.

```{r join}
# !! "unquoting" operator enforces immediate evaluation of the left-side variable inside the function.
# := "glue" operator (frequently used instead of = in rename), allows dynamical mapping of columns' names.

library(dplyr)
library(tibble)
in_names <- ls(pattern = "_sig$")

fc_sam_cold <- data.frame(Name = character())

for (name in in_names) {
  in_obj=as.data.frame(get(name))
  
  x = gsub("_sam_sig", "_S_cold", name)
  new_name = gsub("d1", "", x)
  
  fc = in_obj |>
    rownames_to_column("Name") |>
      select(Name, log2FoldChange) |>
        rename(!!new_name := log2FoldChange)
  
  fc_sam_cold = full_join(fc_sam_cold, fc, by = "Name")
}

rm(list = in_names)
```


Load and process cluster data saved in [07-SAM-only-mfuzzALLsig.qmd](07-SAM-only-mfuzzALLsig.qmd) step.

```{r cluster-data}
load("../rdata-saved/mfuzz_c10_m1.25_samALLsig.RDa")

## Extract assigned cluster number and all membership values
cl_vec   <- c10_m1.25$cluster               # vector
mem_mat  <- c10_m1.25$membership            # matrix

## Check if names are in concert
identical(names(cl_vec), rownames(mem_mat))

## Create data frame
df <- data.frame(
  cluster = cl_vec,                     # 1st column - cluster
  mem_mat,                              # next 10 columns - membership values
  check.names = FALSE                   # allow "1", "2", … as colnames
)

## Rename membership columns
colnames(df)[2:ncol(df)] <- paste0("mem_", seq_len(ncol(mem_mat)))

## Filter, rename and select needed columns

df2 <- df |>
  rownames_to_column(var = "Name") |>
    rename(cluster_S_cold = cluster)

# From cluster-membership table extract only genes with membership >=0.7
df_filt <- df2 |>
  filter(if_any(starts_with("mem_"), ~ . >= 0.7) ) |> # condition on each
    select(Name, cluster_S_cold)
```

Join with DE-analysis results.

```{r de-cluster}
sam_cold = left_join(df_filt, fc_sam_cold, by = "Name", relationship = "one-to-one")

save(sam_cold, file = "../rdata-saved/sam_cold.DE_cluster_table.RDa")
```

Result tables from different tissues and projects are joinded in [14-ALL-projects-genes-cluster-table.qmd](14-ALL-projects-genes-cluster-table.qmd).