---
title: "WGCNA for a554 leaf in cold + regrowth. Part VI - line plots of expresion for modules."
format: html
date: "2026-06-29"
bibliography: ../../../refs-stat-anal.bib
execute:
  fig-format: png
  fig.dpi: 300
---

## Rationale

This script is based on [Examine Expression Profiles](https://bioinformaticsworkbook.org/tutorials/wgcna.html#gsc.tab=0).

The aim is to visualize genes from WGCNA modules.

Built with R version `r getRversion()`

## R Preliminaries

Environment cleaning and set-up

```{r setup, include = FALSE}
#library(pheatmap)
library(DESeq2)
#library(RColorBrewer)
#library(paletteer)
library(limma)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(Mfuzz)
```

## Load data

```{r load}
rm(list = ls())

# set variables

tissue.name = "leaf"

line.name = "a554"

power = 13

# counts after vsd - the set computed for consensus analysis
load('../../../../leaf/rdata-saved/wgcna/vsd-mad.RDa')

in_file = paste0('../rdata-saved/modules4go', power, tissue.name, "-", line.name, '.RDa')

# modules
load(in_file)
```


## Plots for individual replicates

### Prepare data

```{r sort}
coldata_srt = coldata.leaf[order(coldata.leaf$line, coldata.leaf$timepoint),]

vsd.counts_srt <- vsd.counts[, rownames(coldata_srt)]
```

```{r check}
# Check if files are in concert
all(rownames(coldata_srt) %in% colnames(vsd.counts_srt))
all(rownames(coldata_srt) == colnames(vsd.counts_srt))
```

```{r counts modules}
# tables with genes in each module, reverse sorted
counts_genes <- lapply(names(genes_modules), function(name) {
  ids_present <- intersect(genes_modules[[name]], rownames(vsd.counts_srt))
  df <- as.data.frame(vsd.counts_srt[ids_present, , drop = FALSE])
  df$mean_expr <- rowMeans(df)
  df <- df[order(df$mean_expr, decreasing = TRUE), , drop = FALSE]
  df$mean_expr <- NULL
  df
})
names(counts_genes) <- names(genes_modules)

# Make one data frame
counts_df <- bind_rows(counts_genes, .id = "source")

```

### Plot

```{r prepare data}
# modules_of_interest = c("cyan",
#                         "darkgreen",
#                         "darkgrey",
#                         "ivory",
#                         "paleturquoise",
#                         "plum1",
#                         "royalblue",
#                         "turquoise",
#                         "violet")

modules_of_interest = names(counts_genes)

# Pull out list of genes in that module - this step make sense when only a subset of modules are plotted.
# Otherwise it only make a copy of info2
submod = info2 |>
  subset(ModuleColor %in% modules_of_interest)

# Select matching counts
subexpr = counts_df[submod$Probe,]

submod_df = data.frame(subexpr) |>
  rownames_to_column(var = "gene_id") |>
    pivot_longer(-c(gene_id, source)) |>
      mutate(module = source) |>
        select(-source)
```

### Plots for all reps

```{r plot}

out_plot = paste0(tissue.name, power, line.name, "_reps.png")

png(filename = out_plot, width = 2000, height = 16000, res = 300, type = "cairo")

submod_df %>% ggplot(., aes(x=name, y=value, group=gene_id)) +
  geom_line(aes(color = module),
            alpha = 0.2) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90)
  ) +
  facet_grid(rows = vars(module)) +
  labs(x = "sample",
       y = "vst - normalized counts")

dev.off()
```

## Plots for averaged replications

### Averaging and standardising

```{r avg and std}
# varMetadata - description of variables
metadata=data.frame(labelDescription=c("pooled sample", "letter", "tissue", "line", "hour of pooling", "sample number", "day", "replication", "condition", "id", "reads [million]", "batch", "million of reads binned", "chronological time point", "group"), row.names=colnames(coldata_srt))

# metadata
phenoData=new("AnnotatedDataFrame", data=as.data.frame(coldata_srt), varMetadata=metadata)
phenoData

## check
# sampleNames(phenoData)
# varMetadata(phenoData)
# head(pData(phenoData))

# make ExpressionSet
vsd.eset <- ExpressionSet(assayData=vsd.counts_srt, phenoData=phenoData)

# average reps
average = t(avereps(t(exprs(vsd.eset)), ID = vsd.eset$group)) # output is a matrix

# check
dim(average)
dim(vsd.eset)

## make new phenoData
phenoDataAve = phenoData[!duplicated(vsd.eset$group), ]
rownames(phenoDataAve)= phenoDataAve$group

pdata_sorted <- pData(phenoDataAve) |>
  arrange(line, day, time)

# recreate AnnotatedDataFrame:
pData(phenoDataAve) <- pdata_sorted

# sort counts columns
average.srt <- average[, rownames(pData(phenoDataAve))]

all(rownames(pData(phenoDataAve) ) %in% colnames(average.srt))
all(rownames(pData(phenoDataAve) ) == colnames(average.srt))

vsd.eset.ave=ExpressionSet(assayData=average.srt, phenoData=phenoDataAve)

# standardise

vsd.eset.ave.std=standardise(vsd.eset.ave)

# extract expression matrix

counts_ave_std = exprs(vsd.eset.ave.std)

# check
head(counts_ave_std)

# Make reduced coldata for plotting
coldata_ave <- coldata_srt[!duplicated(coldata_srt$group), ]
```

### Split averaged counts by module

```{r counts modules}
# lista tabelek z genami z modułów, posortowane malejąco
counts_genes_ave <- lapply(names(genes_modules), function(name) {
  ids_present <- intersect(genes_modules[[name]], rownames(average))
  df <- as.data.frame(average[ids_present, , drop = FALSE])
  df$mean_expr <- rowMeans(df)
  df <- df[order(df$mean_expr, decreasing = TRUE), , drop = FALSE]
  df$mean_expr <- NULL
  df
})
names(counts_genes_ave) <- names(genes_modules)

# Make one data frame
ave_counts_df <- bind_rows(counts_genes_ave, .id = "source")
```


**The same for standardized data**

```{r counts modules std}
# lista tabelek z genami z modułów, posortowane malejąco
counts_genes_ave.std <- lapply(names(genes_modules), function(name) {
  ids_present <- intersect(genes_modules[[name]], rownames(counts_ave_std))
  df <- as.data.frame(counts_ave_std[ids_present, , drop = FALSE])
  df$mean_expr <- rowMeans(df)
  df <- df[order(df$mean_expr, decreasing = TRUE), , drop = FALSE]
  df$mean_expr <- NULL
  df
})
names(counts_genes_ave.std) <- names(genes_modules)

# Make one data frame
ave_std_counts_df <- bind_rows(counts_genes_ave.std, .id = "source")
```

### Plot

```{r plot ave}
# # modules_of_interest = c("grey60", "violet", "royalblue")
# modules_of_interest = c("darkturquoise", "violet", "royalblue")
# 
# # Pull out list of genes in that module
# submod = info |>
#   subset(ModuleColor %in% modules_of_interest) |>
#     select(Probe, ModuleColor)
# 
# # Select matching counts
# subexpr = ave_counts_df[submod$Probe,]
# 
# submod_df = data.frame(subexpr) |>
#   rownames_to_column(var = "gene_id") |>
#     pivot_longer(-c(gene_id, source)) |>
#       mutate(module = source) |>
#         select(-source)
# 
# 
# png(filename = "test_plot_modules_ave_dkturq.png", width = 2000, height = 1200, res = 300)
# 
# submod_df %>% ggplot(., aes(x=name, y=value, group=gene_id)) +
#   geom_line(aes(color = module),
#             alpha = 0.2) +
#   theme_bw() +
#   theme(
#     axis.text.x = element_text(angle = 90)
#   ) +
#   facet_grid(rows = vars(module)) +
#   labs(x = "sample",
#        y = "vst - normalized counts")
# 
# dev.off()
```

## Plots for averaged reps, splitted by line

```{r line-time}
# Select matching counts
subexpr = ave_counts_df[submod$Probe,]

submod_df = data.frame(subexpr) |>
  rownames_to_column(var = "gene_id") |>
    pivot_longer(names_to = c('line', 'time'), 
                 names_sep = '_', 
                 values_to = 'vst_counts', 
                 -c(gene_id, source)) |>
      mutate(module = source) |>
        select(-source)
```

### Plot

<!-- ```{r} -->
<!-- png(filename = "test_plot_modules_ave_ln-tm.png", width = 2000, height = 1200, res = 300) -->

<!-- submod_df %>% ggplot(., aes(x=time, y=vst_counts, group=gene_id)) + -->
<!--   geom_line(aes(color = module), -->
<!--             alpha = 0.2) + -->
<!--   theme_bw() + -->
<!--   theme( -->
<!--     axis.text.x = element_text(angle = 90) -->
<!--   ) + -->
<!--   facet_grid(rows = vars(module), cols = vars(line)) +  -->
<!--   labs(x = "sample", -->
<!--        y = "vst - normalized counts") -->

<!-- dev.off() -->
<!-- ``` -->


<!-- ```{r} -->
<!-- submod_df2 <- submod_df %>% mutate(time = factor(time, levels = unique(time))) -->

<!-- png(filename = "test_plot_modules_ave_ln-tm2.png", width = 2000, height = 1200, res = 300) -->

<!-- submod_df2 %>% -->
<!--   ggplot(aes(x = time, y = vst_counts, group = gene_id, color = module)) + -->
<!--   geom_line(alpha = 0.2) + -->
<!--   theme_bw() + -->
<!--   theme(axis.text.x = element_text(angle = 90)) + -->
<!--   facet_grid(rows = vars(module), cols = vars(line)) + -->
<!--   labs(x = "sample", y = "vst - normalized counts") -->
<!-- dev.off() -->
<!-- ``` -->

```{r}

submod_df2 <- submod_df %>% mutate(time = factor(time, levels = unique(time)))

out_plot2 = paste0(tissue.name, power, line.name, "_ave.png")

# Best version, uses module names as colors
png(filename = out_plot2, width = 2000, height = 16000, res = 300, type = "cairo")
submod_df2 %>%
  ggplot(aes(x = time, y = vst_counts, group = gene_id, color = module)) +
  geom_line(alpha = 0.2) +
  scale_color_identity() +   # use module value as color names
  facet_grid(rows = vars(module), cols = vars(line)) +
  theme_dark() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "sample", y = "vst - normalized counts")
dev.off()
```


**The same for averaged and standardized reps**

```{r line-time}
# Select matching counts
subexpr_std = ave_std_counts_df[submod$Probe,]

submod_df_std = data.frame(subexpr_std) |>
  rownames_to_column(var = "gene_id") |>
    pivot_longer(names_to = c('line', 'time'), 
                 names_sep = '_', 
                 values_to = 'vst_counts', 
                 -c(gene_id, source)) |>
      mutate(module = source) |>
        select(-source)
```

```{r}

submod_df2_std <- submod_df_std %>% mutate(time = factor(time, levels = unique(time)))

out_plot3light = paste0(tissue.name, power, line.name, "_ave_std_light.png")
out_plot3dark = paste0(tissue.name, power, line.name, "_ave_std_dark.png")

# Best version, uses module names as colors

# LIGHT background

png(filename = out_plot3light, width = 2000, height = 16000, res = 300) # LIGHT
submod_df2_std %>%
  ggplot(aes(x = time, y = vst_counts, group = gene_id, color = module)) +
  geom_line(alpha = 0.2) +
  scale_color_identity() +   # use module value as color names
  facet_grid(rows = vars(module), cols = vars(line)) +
  theme_bw() + # LIGHT
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "sample", y = "vst, std counts") +
  geom_vline(xintercept = 6.5, color = "white", linewidth = 3) # WHITE
dev.off()

# DARK background

png(filename = out_plot3dark, width = 2000, height = 16000, res = 300) # DARK
submod_df2_std %>%
  ggplot(aes(x = time, y = vst_counts, group = gene_id, color = module)) +
  geom_line(alpha = 0.2) +
  scale_color_identity() +   # use module value as color names
  facet_grid(rows = vars(module), cols = vars(line)) +
  theme_dark() + # better for light module colors # DARK
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "sample", y = "vst, std counts") +
  geom_vline(xintercept = 6.5, color = "#7f7f7f", linewidth = 3) # DARK
dev.off()
```

## References
