---
title: "WGCNA for a554 leaf in cold + regrowth. Part III - data export for further analysis."
format: html
date: "2026-06-26"
execute:
  fig-format: png
  fig.dpi: 300
---

Built with R version `r getRversion()`

This script continues [02-a554-MAD-wgcna-leaf-cold-2network.qmd](02-a554-MAD-wgcna-leaf-cold-2network.qmd).

And is based on "Tutorial for the WGCNA package for R: I. Network analysis of liver expression data in female mice 3. Relating modules to external information and identifying important genes".

# Data preparation

## Environment cleaning and set-up

```{r setup, include = FALSE}
# load required packages
# rm(list = ls())
library(WGCNA)

# The following setting is important, do not omit.
options(stringsAsFactors = FALSE)
# Allow multi-threading within WGCNA.
# Caution: skip this line if you run RStudio or other third-party R environments.
# See note above.
enableWGCNAThreads()

library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r load}

rm(list = ls())

# set variables

tissue.name = "leaf"

line.name = "a554"

power = 13

# read objects

in_file = paste0('../rdata-saved/networkConstruction', power, tissue.name, "-", line.name, '.RDa')

load(in_file)

in_file2 = paste0("../rdata-saved/", tissue.name, "-", line.name, "-dataInput.RData")

load(in_file2)
```

# Create objects for GO-enrichment analysis

```{r files for GO}

# All genes after MAD-filtering, it will be population for GO-enrichment
mad.genes = names(datExpr)

# Genes in modules
info2 = data.frame(Probe = names(datExpr),
                   ModuleColor = moduleColors) |> 
        filter(moduleColors != "grey")

# Create list
genes_modules <- split(info2$Probe, info2$ModuleColor)# 

out_file = paste0('../rdata-saved/modules4go', power, tissue.name, "-", line.name, '.RDa')

save(genes_modules, info2, mad.genes, file = out_file)
```

:::{.callout-note}
Other means to export only gene IDs and module color - [section 'Relate Module (cluster) Assignments to Treatment Groups'](https://bioinformaticsworkbook.org/tutorials/wgcna.html#gsc.tab=0). And for Cytoscape: 'Generate and Export Networks'.
:::
