---
title: "WGCNA for a554 leaf in cold + regrowth. Part II - network construction."
format: html
date: "2026-06-24"
execute:
  fig-format: png
  fig.dpi: 300
---

Built with R version `r getRversion()`

This script continues [02-a554-MAD-wgcna-leaf-cold-1data.qmd](02-a554-MAD-wgcna-leaf-cold-1data.qmd).

# Data preparation

## Environment cleaning and set-up

```{r setup, include = FALSE}
# load required packages
#rm(list = ls())
library(WGCNA)

# The following setting is important, do not omit.
options(stringsAsFactors = FALSE)
# Allow multi-threading within WGCNA.
enableWGCNAThreads()
```

```{r suffix}
rm(list = ls())

tissue.name = "leaf"

line.name = "a554"

```

```{r load}
in_file = paste0("../rdata-saved/", tissue.name, "-", line.name, "-dataInput.RData")

# Load the data saved in the first part
load(in_file)

```

## Step-by-step network construction and module detection

### Choosing the soft-thresholding power: analysis of network topology

The three different network types are coded and quoted in `pickSoftThreshold` function.
Likewise, there is `bicor` version.

:::{.callout-important}
According to FAQ:
> We strongly recommend using the argument maxPOutliers = 0.05 or 0.10 whenever the biweight midcorrelation is used
:::

```{r thresholding powers, cache=TRUE}
# Choose a set of soft-thresholding powers
# powers = c(seq(10, 30, by=1))
powers = c(seq(5, 15, by=1))
# # Initialize a list to hold the results of scale-free analysis
# powerTables = vector(mode = "list", length = nSets)

# Call the network topology analysis function for each set in turn
# for (set in 1:nSets)
sft = pickSoftThreshold(datExpr, 
                               powerVector=powers,
                               verbose = 2, 
                               networkType = "signed hybrid", 
                               corFnc = 'bicor', 
                               corOptions = list(maxPOutliers = 0.05),
                               moreNetworkConcepts = TRUE)
# additional measures ("moreNetworkConcepts") described in https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1000117
# Density : close to 1, all genes are strongly correlated, ~ 0.5 more ambigous correlations
# Centralization: close to 1, one gene strongly correlated with all others which in turn have negative correlations
# Heterogeneity: can be used to measure variation of connectivity of genes.

# Optionally
# networkType = "signed", corFnc = 'bicor'
# networkType = "signed hybrid", corFnc = 'bicor'

collectGarbage()
```

Plotting results
```{r soft-thr, fig.show='hold', cache=TRUE, fig.width=12, fig.height=7, dpi=300, dev='png'}
# sizeGrWindow(9, 5) # used in tutorial but here omitted
par(mfrow = c(2,2))
cex1 = 0.7
# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
  xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
  main = paste("Scale independence"));
  text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
  labels=powers, cex=cex1, col="red")
# this line corresponds to using an R^2 cut-off of h
abline(h=0.8,col="blue")

# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
  xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
  main = paste("Mean connectivity"))
  text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1, col="red")
abline(h=50,col="blue")
  
# Median connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,6],
  xlab="Soft Threshold (power)",ylab="Median connectivity", type="n",
  main = paste("Median connectivity"))
  text(sft$fitIndices[,1], sft$fitIndices[,6], labels=powers, cex=cex1, col="red")
abline(h=10,col="blue")
  
# Max connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,7],
  xlab="Soft Threshold (power)",ylab="Max connectivity", type="n",
  main = paste("Max connectivity"))
  text(sft$fitIndices[,1], sft$fitIndices[,7], labels=powers, cex=cex1, col="red")
```

:::{.callout-important}
From [thread](https://support.bioconductor.org/p/121622/#121653).
Peter Langfelder (the package developer) regarding "mean connectivity" advises:
> I aim for mean connectivity around 100 or less (typically between 30-50); median connectivity usually ends up around 10.

Also, the FAQ states that for "signed hybrid" power should by < 15.
Using "signed hybrid", potentially power 13 is good, based on "SFT.R.sq", "mean.k" and "median.k", averaged across samples.
However mean connectivity is notably lower than advised by author.

:::

```{r save-intermediate}
out_file = paste0("../rdata-saved/", tissue.name, "-", line.name, "-before-adj.RDa")

save.image(out_file)
```

<!-- ################ START of code executed externally ########################## -->


### Calculation of network adjacencies

:::{.callout-note}
As they are computationally demanding the adjacencies and TOMsimilarity calculations were moved to external R file for batch running [TOMs_powers1line.R](TOMs_powers1line.R).
:::

### Calculation of topological overlap

:::{.callout-important}
`TOMtype` is different than `networkType`, so using `networkType = "signed"` does not imply using `TOMtype = "signed"`!!! [post](https://support.bioconductor.org/p/73814/#73894)
:::

<!-- ################ END of code executed externally ########################## -->

```{r load TOM}
rm(list = ls())

power = 13

line.name = "a554"

tissue.name = "leaf"

in_file = paste0('../rdata-saved/power', power, tissue.name, "-", line.name, '.RDa')

load(in_file)

in_file2 = paste0("../rdata-saved/", tissue.name, "-", line.name, "-before-adj.RDa")

load(in_file2)
```


### Clustering and module identification

Calculate the dissimilarity:
```{r}
dissTOM = 1-TOM
```

We now use hierarchical clustering to produce a hierarchical clustering tree (dendrogram) of genes.

```{r dendro, fig.width=12, fig.height=9}
# Call the hierarchical clustering function
geneTree = hclust(as.dist(dissTOM), method = "average")
# Plot the resulting clustering tree (dendrogram)
# sizeGrWindow(12,9) # For interactive work only.
plot(geneTree, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity",
  labels = FALSE, hang = 0.04)
```


In the clustering tree (dendrogram), each leaf, that is a short vertical line, corresponds to a gene. Branches of the dendrogram group together densely interconnected, highly co-expressed genes. Module identification amounts to the identification of individual branches (”cutting the branches off the dendrogram”). There are several methods for branch cutting; our standard method is the Dynamic Tree Cut from the package `dynamicTreeCut`. The next snippet of code illustrates its use.

```{r modules}
# We like large modules, so we set the minimum module size relatively high:
minModuleSize = 30
# Module identification using dynamic tree cut:
dynamicMods = cutreeDynamic(dendro = geneTree, 
                            distM = dissTOM,
                            deepSplit = 2, # The higher the value the more and smaller clusters will be produced. 
                            cutHeight = 0.995,
                            pamRespectsDendro = FALSE, # ? default is TRUE but in manual FALSE is given
                            minClusterSize = minModuleSize)

# quick summary of the module detection
table(dynamicMods)
```

The Dynamic Tree Cut returned 117! proper modules. The label 0 (grey) is
reserved for genes not assigned to any of the modules. The following code plots the consensus gene dendrogram
together with the preliminary module colors:

```{r consensus dendrogram, fig.width=8, fig.height=6, dpi=300, dev='png'}
# Convert numeric lables into colors
dynamicColors = labels2colors(dynamicMods)
table(dynamicColors)
# Plot the dendrogram and colors underneath

plotDendroAndColors(geneTree, 
                    dynamicColors, 
                    "Dynamic Tree Cut",
                    dendroLabels = FALSE, 
                    hang = 0.03,
                    addGuide = TRUE, 
                    guideHang = 0.05,
                    main = "Gene dendrogram and module colors")
```

### Merging of modules whose expression profiles are very similar

The Dynamic Tree Cut may identify modules whose expression profiles are very similar. It may be prudent
to merge such modules since their genes are highly co-expressed. To quantify co-expression similarity of entire
modules, we calculate their eigengenes (MEs) and cluster them on their consensus correlation, that is the minimum
correlation across the two sets:

```{r consensus module eigengenes, fig.width=7, fig.height=6, dpi=300, dev='png'}
# Calculate eigengenes
MEList = moduleEigengenes(datExpr, colors = dynamicColors)
MEs = MEList$eigengenes
# Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs)
# Cluster module eigengenes
METree = hclust(as.dist(MEDiss), method = "average");
# Plot the result
# sizeGrWindow(7, 6) # For interactive work only.
plot(METree, main = "Clustering of module eigengenes",
  xlab = "", sub = "", cex = 0.4)
# We choose a height cut of 0.25, corresponding to correlation of 0.75, to merge.
## In fact determined after plotting, here included to avoid error.
MEDissThres = 0.25
# Plot the cut line into the dendrogram
abline(h=MEDissThres, col = "red")
```

Some clusters can be merged.

```{r consensus merge modules}
# Call an automatic merging function
merge = mergeCloseModules(datExpr, dynamicColors, cutHeight = MEDissThres, verbose = 3)
# The merged module colors
mergedColors = merge$colors

# 3. TUTAJ WKLEJASZ KOD NAPRAWIAJĄCY ETYKIETY (Zamiast starego match())
# =====================================================================
customLevels = c("grey", setdiff(unique(mergedColors), "grey"))
moduleLabels = as.numeric(factor(mergedColors, levels = customLevels)) - 1

# Eigengenes of the new merged modules:
mergedMEs = merge$newMEs
```

To see what the merging did to our module colors, we plot the gene dendrogram again, with the original and merged module colors underneath:

```{r dendro-compare, fig.width=12, fig.height=9}
# sizeGrWindow(12, 9) # For interactive work only.
plotDendroAndColors(geneTree, 
                    cbind(dynamicColors, mergedColors),
                    c("Dynamic Tree Cut", "Merged dynamic"),
                    dendroLabels = FALSE, 
                    hang = 0.03,
                    addGuide = TRUE, 
                    guideHang = 0.05)
```


```{r merge stat}
table(mergedColors)
```

After merge there are 28 modules.

In the subsequent analysis, we will use the merged module colors in `mergedColors`. We save the relevant variables:

```{r}
# Rename to moduleColors
moduleColors = mergedColors

MEs = mergedMEs

# Save module colors and labels for use in subsequent parts
out_file2 = paste0('../rdata-saved/networkConstruction', power, tissue.name, "-", line.name, '.RDa')

save(MEs, moduleLabels, moduleColors, geneTree, file = out_file2)
```

:::{.callout-note}
Proceed to [02-a554-MAD-wgcna-leaf-cold-3export.qmd](02-a554-MAD-wgcna-leaf-cold-3export.qmd).
:::


# References