---
title: "WGCNA for a554 leaf in cold + regrowth. Part I - data input. Filtering on MAD".
format: html
date: "2026-06-24"
execute:
  fig-format: png
  fig.dpi: 300
---

Built with R version `r getRversion()`

# Rationale

6 timepoints are available per line per variant (cold / control). There are 3 biological replications so in all we have 18 samples per line per variant. "time of day" (three letters) * "day" (1, 2) * "biological replication" (1 to 3) $3 * 2 * 3 = 18$ samples. Times 2 variants (temperature) gives 36 samples per line.
The sample size of joined temperature variants is sufficient according to FAQ ('at least 30').

The aim is to perform WGCNA - separate per line.


:::{.callout-important}
The results of per line analysis will be used as input to `modulePreservation` function.
So for each line the *same set of genes* is used - the one filtered in [consensus analysis](../../../../leaf/scripts/02-MAD-wgcna-leaf-cold-line4-1data.qmd).

In the linked script two filtering steps were performed:
1. At least 10 counts in at least 90% samples.
2. MAD above 1st quartile.
:::

# Data preparation

## Environment cleaning and set-up

```{r setup, include = FALSE}
# load required packages
library(GenomicRanges)
library(WGCNA)
library(dplyr)
library(DESeq2)
```


```{r instal-packages}
#| eval: false
#| output: false
#| message: false
# Clean working directory, if needed
rm(list=ls())

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("impute")
BiocManager::install("preprocessCore")
BiocManager::install("GO.db")
BiocManager::install("AnnotationDbi")

install.packages("WGCNA")
```

## Read-in data

```{r read data, cache=TRUE}
load("../../../../../input-data/rdata/counts_and_metadata.RDa")

# previously prepared range data
load("../../../../../input-data/rdata/ranges.RDa")

# transform Factors to chr/int where appropriate
coldata2 = coldata |> mutate(across(letter : line, as.character), 
                             condition = as.character(condition), 
                             across(time : rep, as.integer))

# Overwrite original object to keep further code unchanged
coldata = coldata2

# Check if files are in concert
all(rownames(coldata) %in% colnames(cts))
all(rownames(coldata) == colnames(cts))
```

## Subset leaf and a554 data


```{r suffix}

tissue.name = "leaf"

line.name = "a554"

```


Subsetting is done at the level of raw data (counts and coldata).
```{r subset}
# subset coldata
coldata.leaf = coldata |> filter(tissue == tissue.name & line == line.name)

# check
dim(coldata.leaf)

# vector of sample names to keep - needed to select columns from count file
keep.leaf=rownames(coldata.leaf)

# check - only samples named "la5..." should remain
keep.leaf
length(keep.leaf)

# subsetting counts
cts.leaf=subset(cts, select=keep.leaf)

# check
dim(cts.leaf)
colnames(cts.leaf)

# Check if counts and coldata are in concert
all(rownames(coldata.leaf) %in% colnames(cts.leaf))
all(rownames(coldata.leaf) == colnames(cts.leaf))

keep = c("coldata.leaf", "cts.leaf", "gzakresy", "tissue.name", "line.name")

rm(list = setdiff(ls(), keep))
```

## Data preparation

DESeq data set must be created to perform VST.

```{r factors}
#| output: false
#| message: false

## Prepare grouping variable for use in model in dds

# add variable describing chronology of samples
coldata.leaf$timepoint = coldata.leaf$time + (coldata.leaf$day - 1) * 3

# Adding grouping variable to easily analyse all comparisons needed
coldata.leaf$group <- factor(paste0(coldata.leaf$line, "_", coldata.leaf$timepoint))

# check
str(coldata.leaf)
```

## Create DEseq Data Set

```{r dds, cache=TRUE}
#| output: false
#| message: false

# create dataset, design included only for command completeness, different will be used in DE analysis
dds.leaf = DESeqDataSetFromMatrix(countData = cts.leaf, colData = coldata.leaf, rowRanges = gzakresy, design = ~ group)
```

## Pre-filtering

:::{.callout-important}
Using genes kept in consensus analysis [consensus analysis](../../../../leaf/scripts/02-MAD-wgcna-leaf-cold-line4-1data.qmd).
:::

```{r genes to keep}
load('../../../../leaf/rdata-saved/wgcna/genes2keep.RDa')

keep.consensus2 <- names(keep.consensus)[keep.consensus]
```

```{r filter}

dds.leaf.f <- dds.leaf[keep.consensus2,]

# check
dim(dds.leaf.f)

# what percent of genes remained?
100*(dim(dds.leaf.f)[1]/dim(dds.leaf)[1])

# [1] 13492    36
# [1] 30.45392
```

## Normalization

According to [post](https://www.biostars.org/p/459013/#459016) `varianceStabilizingTransformation` not `vst` is used. Design is provided here only because it is required in `dds` object. It makes no sense using `blind=FALSE` here.

```{r normalize, cache=TRUE}
vsd.leaf = varianceStabilizingTransformation(dds.leaf.f, blind=TRUE)

rm(list = setdiff(ls(), c('vsd.leaf', 'coldata.leaf', 'line.name', 'tissue.name')))
```

## Extract normalized counts

```{r extract normalized counts}
vsd.counts = assay(vsd.leaf)

coldata = colData(vsd.leaf)

# check
dim(vsd.counts)
dim(coldata)
```

# WGCNA analysis

## Data input, cleaning & preprocessing


```{r}
#| output: false

options(stringsAsFactors = FALSE) # Important option, do not omit!
enableWGCNAThreads()
```

## Reformat data

```{r format data}
dim(vsd.counts)

# the data contains additional variables - next code removes them and transposes (very important!) the data
datExpr = as.data.frame(t(vsd.counts))

# check row and col names
head(names(datExpr))
head(rownames(datExpr))

```

## Checking data for excessive missing values and identification of outlier samples

```{r check missing}
# Check the data for missing values
gsg = goodSamplesGenes(datExpr, verbose = 3)
gsg$allOK
```

## Clustering of samples to see if there are obvious outliers

```{r fig.width=12, fig.height=9}
sampleTree = hclust(dist(datExpr), method = "average")

pdf(file = "sampleClustering.pdf", width = 12, height = 9);
par(cex = 0.6);
par(mar = c(0,4,2,0))
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5,
cex.axis = 1.5, cex.main = 2)
abline(h = 15, col = "red")
dev.off()
```

There are no outliers, samples clusters by temperature but not perfectly.

# Loading category data

It is essentially coldata information. May be used in plots.

```{r cats}

# select interesting columne
datTraits = coldata.leaf[, (names(coldata.leaf) %in% c('time', 'day', 'rep'))]

```

We now have the expression data in the variable `datExpr`, and the corresponding clinical traits in the variable
`datTraits`. Before we continue with network construction and module detection, we visualize how the clinical traits
relate to the sample dendrogram.

```{r fig.width=12, fig.height=9}
# Re-cluster samples
sampleTree2 = hclust(dist(datExpr), method = "average")
# Convert traits to a color representation: white means low, red means high, grey means missing entry
traitColors = numbers2colors(datTraits, signed = FALSE)
# Plot the sample dendrogram and the colors underneath.
plotDendroAndColors(sampleTree2, traitColors,
groupLabels = names(datTraits),
main = "Sample dendrogram and trait heatmap (a554)")
```

```{r save}

out_file = paste0("../rdata-saved/", tissue.name, "-", line.name, "-dataInput.RData")

save(datExpr, datTraits, file = out_file)

```

:::{.callout-note}
Proceed to [02-a554-MAD-wgcna-leaf-cold-2network.qmd](02-a554-MAD-wgcna-leaf-cold-2network.qmd).
:::


# References