---
title: "WGCNA for a554 leaf in cold + regrowth. Part IV - cytoscape networks for top nodes"
format: html
date: "2026-06-26"
bibliography: ../../../refs-stat-anal.bib
execute:
  fig-format: png
  fig.dpi: 300
---

## Rationale
The aim of this script is to filter network nodes to generate cytoscape networks of similar size with top hubs.

Built with R version `r getRversion()`

## R Preliminaries

Environment cleaning and set-up

```{r setup, include = FALSE}
# load required packages
# rm(list = ls())
library(WGCNA)

# The following setting is important, do not omit.
options(stringsAsFactors = FALSE)
# Allow multi-threading within WGCNA.
enableWGCNAThreads()
# 
# library(dplyr)
# library(tidyr)
# library(ggplot2)

# load package
library(RCy3)

# check connection to Cytoscape
cytoscapePing()
```

```{r load data}
rm(list = ls())

# set variables

tissue.name = "leaf"

line.name = "a554"

power = 13

in_file = paste0('../rdata-saved/networkConstruction', power, tissue.name, "-", line.name, '.RDa')

load(in_file) # datExpr (for gene names), TOM and moduleColors

in_file2 = paste0('../rdata-saved/power', power, tissue.name, "-", line.name, '.RDa')

load(in_file2)

in_file3 = paste0("../rdata-saved/", tissue.name, "-", line.name, "-dataInput.RData")

load(in_file3)

# Read in the annotation file
annot = read.table("../../../../../input-data/go-uniprot/NAM-UP-anno", sep = "\t", header = FALSE)
```

# Create network files for top 100 hubs

```{r cyto loop}
## Method to gen all modules

# set variables
x = unique((moduleColors))

modules = x[x != 'grey']

thresh = 0.02

probes = colnames(datExpr) # select colnames from one of the sets

# 5. Określ liczbę top-hubów (podobnie jak w oryginalnym skrypcie)
nTop = 100

for (module in modules) {

# 2. Wyodrębnij sondy należące do tego modułu
inModule = (moduleColors == module)
modProbes = probes[inModule]

# 3. Wyciągnij Topological Overlap z wyliczonej macierzy CONSENSUS TOM
# Zamiast TOM używamy consTOM wyliczonej podczas budowy sieci consensus
modTOM = TOM[inModule, inModule]
dimnames(modTOM) = list(modProbes, modProbes)

# 4. Oblicz łączność węzłów wewnątrz modułu na podstawie consTOM
# Suma wierszy (rowSums) w topologicznym overlapie świetnie odzwierciedla wewnątrzmodułową łączność 
IMConn = rowSums(modTOM)

# 6. Stwórz wektor logiczny "top"
top = (rank(-IMConn) <= nTop)

# 7. Wyciągnij symbole genów dla całego modułu (na podstawie adnotacji)
modProts = annot$V2[match(modProbes, annot$V1)]

# 8. Wyeksportuj odpowiednio skróconą sieć do Cytoscape
cyt = exportNetworkToCytoscape(modTOM[top, top],
  edgeFile = paste0("../results/wgcna/network/Consensus-CytoscapeInput-edges-", module, "-top", nTop, ".txt"),
  nodeFile = paste0("../results/wgcna/network/Consensus-CytoscapeInput-nodes-", module, "-top", nTop, ".txt"),
  weighted = TRUE,
  threshold = thresh,
  # Upewnij się, że atrybuty i nazwy też filtrujesz wektorem [top]
  nodeNames = modProbes[top],
  altNodeNames = modProts[top],
  nodeAttr = moduleColors[inModule][top] 
)

}
```

# Draw networks for top 100 hubs

:::{.callout-note}
Networks were meant to be colored according to module name, however only basic colors are recognized by cytoscape.

The network generation require tuning - eg. node size by degree is not a good idea or other scale is needed.
:::

```{r create network each}

# set variables
x = unique((moduleColors))

modules = x[x != 'grey']

for (module in modules){
# Set input file names
  
input_nodes=paste0("../results/wgcna/network/Consensus-CytoscapeInput-nodes-", module, "-top", nTop, ".txt")

input_edges=paste0("../results/wgcna/network/Consensus-CytoscapeInput-edges-", module, "-top", nTop, ".txt")

## extract color name

# ncolor = gsub("-.*", "", name)

# Read node and edge files

node_info = read.table(input_nodes, header = TRUE, sep = "\t")
colnames(node_info)[3] = "module"

edge_info = read.table(input_edges, header = TRUE, sep = "\t")

# Create network

createNetworkFromDataFrames(
  title = module,
  collection = "leaf, ctrl, wgcna",
  nodes = node_info,
  node.id.list = "nodeName",
  source.id.list = "fromNode",
  target.id.list = "toNode",
  edges=edge_info
)

## change node shape

setNodeShapeDefault('ELLIPSE')

## run "Analyze network" to get degree values
analyzeNetwork()

### set node shape mapping

setNodeSizeMapping(
  table.column = "Degree",
  table.column.values = c(1, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100),
  sizes = c(20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70),
  mapping.type = "c"
)

## set edge mapping according to "weight" from WGCNA

setEdgeLineWidthMapping(table.column = 'weight',
                        widths = c(1, 10))

## set module color
### Passtrought mapping - works fine except "darkgrey"

setNodeColorMapping(table.column = "module",
                   mapping.type = "p")

## add annotation table

loadTableData(annot, data.key.column="V1")

}

out_file = paste0("../results/wgcna/network/cold-top", nTop , "power", power, tissue.name, "-", line.name)

saveSession(out_file) #.cys
```