---
title: "Prepare network files for analysis in Cytoscape. Arabidopsis STRING network as basis."
format: html
date: "2026-02-24"
bibliography: ../../../refs-stat-anal.bib
execute:
  fig-format: png
  fig.dpi: 300
---

The aim of this code is to overlay maize orthologues on STRING network of *Arabidopsis thaliana*.
Orthologues of *Arabidopsis thaliana* genes annotated with "Rhythmic process" (GO:0048511) GO term were retrieved from UniProt. The script uses clustering results and String network data [@szklarczykSTRINGDatabase20232023] for Cytoscape analysis and visualization [@clineIntegrationBiologicalNetworks2007]. Files are prepared separately for each cluster using 0.7 membership cutoff.
This script uses GO-annotation and associated files from UniProt, prepared in [00-test-filter.qmd](../../../input-data/go-uniprot/00-test-filter.qmd) and [00-GO-and-mapping-UniProt.qmd](../../../input-data/go-uniprot/00-GO-and-mapping-UniProt.qmd) and intermediate files from [10-NAMzUP-leaf-networks-cytoscape-loop.qmd](10-NAMzUP-leaf-networks-cytoscape-loop.qmd).
Here option 1) using NAM - GO is realized

Built with R version `r getRversion()`

## R Preliminaries

Environment cleaning and set-up

```{r packages}
#| eval: false
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Mfuzz")
BiocManager::install("marray")
BiocManager::install("limma")
BiocManager::install("RCy3")
BiocManager::install("biomaRt")
```

```{r}
#| output: false

# Clean working directory
rm(list=ls())
```

## Load data

`c13_m1.35` contains data for clusters.


## Prepare and export cluster data

```{bash net-dir}
rm -r ../results/network-rhythm-at-string/
mkdir ../results/network-rhythm-at-string/
```

## Filter network data with significant genes and add orthology

`full-network-At.gz` file contains *edge data* for all interactions. *node data* is needed, `mart-all-gramene_up-zea` contains most data

```{bash rhythm net}
# retrieve subnetwork for At "rhythmic" proteins with orthologues in maize
## protein can be either "source" or "target"
cd ../../../input-data/at-ortho_rhythmic-process/
cut -f7 -d $'\t' mart-all-gramene_up-zea | sort -u > at_UP_ortho_zm
zgrep -wf at_UP_ortho_zm full-network-At.gz > UP_At-rhythm_net

## interactions for clock only
awk '
    NR==FNR {           # Czytamy pierwszy plik (at_UP_ortho_zm )
        keep[$1] = 1;       # budujemy tablicę asocjacyjną – klucz = wartość
       next;             # przejdź do kolejnego rekordu (nie analizujemy dalej)
    }
    $1 in keep && $2 in keep {         # Drugi plik (klastry); sprawdzamy kolumnę 1 i 2
        print               # wypisujemy cały wiersz, jeśli wartość jest w tablicy
    }
' at_UP_ortho_zm  UP_At-rhythm_net > UP_At-rhythm_BothProt_net

## interactions where source node is a clock gene
awk '
    NR==FNR {           # Czytamy pierwszy plik (at_UP_ortho_zm )
        keep[$1] = 1;       # budujemy tablicę asocjacyjną – klucz = wartość
       next;             # przejdź do kolejnego rekordu (nie analizujemy dalej)
    }
    $1 in keep {         # Drugi plik (klastry); sprawdzamy kolumnę 1 i 2
        print               # wypisujemy cały wiersz, jeśli wartość jest w tablicy
    }
' at_UP_ortho_zm  UP_At-rhythm_net > UP_At-rhythm_SourceProt_net
```

```{r sigs}
# This chunk is specific for experiment
# List of all significant genes, those kept in clusters (surviving the 0.7 membership cutoff) are not sufficient as target node are allowed to be anything.

load("../rdata-saved/sigs-leaf1_all.RDa")

library(tibble)

object_names <- ls(pattern = "sig$")

for (name in object_names) {
  out_name <- paste0(gsub("_sig", "", name), "_out")
  result <- as.data.frame(get(name)) |> rownames_to_column("Name")
  assign(out_name, result, envir = .GlobalEnv)
}

rm(list=ls(pattern = "sig$")) # remove intermediate objects

# join all and extract only gene names without duplicates

object_names <- mget(ls(pattern = "out$"))

library(dplyr)

all_sigs <- bind_rows(object_names) |> 
  select(Name) |>
    distinct()

rm(list=ls(pattern = "out$")) # remove intermediate objects
```

```{r edge info}
# read network file - Source = clock, Target = anything
edge.info = read.table("../../../input-data/at-ortho_rhythmic-process/UP_At-rhythm_SourceProt_net", header=FALSE, stringsAsFactors = FALSE, sep=' ', quote = "")

colnames(edge.info)=c("up_at1", "up_at2", "neighborhood", "neighborhood_transferred", "fusion", "cooccurence", "homology", "coexpression", "coexpression_transferred", "experiments", "experiments_transferred", "database", "database_transferred", "textmining", "textmining_transferred", "combined_score")

# The resulting file is exactly as needed for RCy3
```

```{r target node info}
### Retrieve orthologs in Zea mays ###
# this info is available for clock proteins but not for all target proteins

library(biomaRt)
# Load database

#---------- --------------------------- get appropriate parameters -----------------------------#
# # identify name of plant database
listEnsemblGenomes()

# select database for plants
ensembl_plants <- useEnsemblGenomes(biomart = "plants_mart")
# what info for Arabidopsis (we use network for this species) is available?
searchDatasets(ensembl_plants, pattern = "Arabidopsis")
#-----------------------------------------------------------------------------------------------#

# select appropriate database
ensembl_arabidopsis <- useEnsemblGenomes(
  biomart = "plants_mart",
  dataset = "athaliana_eg_gene"
)

# Build query

# what filters are available?
filters <- listFilters(ensembl_arabidopsis)

# what can be extracted?
attributes <- listAttributes(ensembl_arabidopsis)

## which one are for maize
attr_zea = searchAttributes(ensembl_arabidopsis, pattern = "Zea")
# podzbior <- df %>% filter(grepl("^k", tekst)) # alternatiuve command

# extract IDs of target node (Arabidopsis UPs)
targets <- unique(net$up_at2)

# try direct function for homologs
#at2zm = getHomologs(targets, arabidopsis_thaliana, zea_mays)
# the function is listed in manual but tR can't find it (?)

# The procedure is 3-step
## 1. retrieve ortholog info
## 2. retrieve mapping between ensembl and uniorot IDs
## 3. join the two tables to get uniprot IDs and orthologs
## 4. rbind all three files

## 1.
# As there is no info. what IDs in STRING are I'll retrieve orthologs using three filters
# generic function
# "UniProt Gene ID"
at2zm_up_gn_id = getBM(
  attributes = c("ensembl_gene_id", 
                 "zmays_eg_homolog_ensembl_gene", 
                 "zmays_eg_homolog_perc_id",
                 "zmays_eg_homolog_perc_id_r1",
                 "zmays_eg_homolog_orthology_confidence",
                 "description"),
  filters = "uniprot_gn_id",
  values = targets,
  mart = ensembl_arabidopsis,
  uniqueRows = TRUE
)

# SwissProt ID
at2zm_swiss = getBM(
  attributes = c("ensembl_gene_id", 
                 "zmays_eg_homolog_ensembl_gene", 
                 "zmays_eg_homolog_perc_id",
                 "zmays_eg_homolog_perc_id_r1",
                 "zmays_eg_homolog_orthology_confidence",
                 "description"),
  filters = "uniprotswissprot",
  values = targets,
  mart = ensembl_arabidopsis,
  uniqueRows = TRUE
)

# TrEMBL ID
at2zm_trembl = getBM(
  attributes = c("ensembl_gene_id", 
                 "zmays_eg_homolog_ensembl_gene", 
                 "zmays_eg_homolog_perc_id",
                 "zmays_eg_homolog_perc_id_r1",
                 "zmays_eg_homolog_orthology_confidence",
                 "description"),
  filters = "uniprotsptrembl",
  values = targets,
  mart = ensembl_arabidopsis,
  uniqueRows = TRUE
)

## 2.
# ID-mapping
# "UniProt Gene ID"
at2zm_up_gn_id_MAP = getBM(
  attributes = c("ensembl_gene_id", 
                 "uniprot_gn_id"),
  filters = "uniprot_gn_id",
  values = targets,
  mart = ensembl_arabidopsis,
  uniqueRows = TRUE
)

# SwissProt ID
at2zm_swiss_MAP = getBM(
  attributes = c("ensembl_gene_id", 
                 "uniprotswissprot"),
  filters = "uniprotswissprot",
  values = targets,
  mart = ensembl_arabidopsis,
  uniqueRows = TRUE
)

# TrEMBL ID
at2zm_trembl_MAP = getBM(
  attributes = c("ensembl_gene_id", 
                 "uniprotsptrembl"),
  filters = "uniprotsptrembl",
  values = targets,
  mart = ensembl_arabidopsis,
  uniqueRows = TRUE
)


## 3.
# join

# "UniProt Gene ID"
set2rbind_up_gn_id = inner_join(at2zm_up_gn_id_MAP, at2zm_up_gn_id, by = "ensembl_gene_id", relationship="many-to-many") |>
  rename(up_at = uniprot_gn_id)

# SwissProt ID
set2rbind_swiss = inner_join(at2zm_swiss_MAP, at2zm_swiss, by = "ensembl_gene_id", relationship="many-to-many") |>
  rename(up_at = uniprotswissprot)

# TrEMBL ID
set2rbind_trembl = inner_join(at2zm_trembl_MAP, at2zm_trembl, by = "ensembl_gene_id", relationship="many-to-many") |>
  rename(up_at = uniprotsptrembl)

## 4.
# rbind

object_names <- mget(ls(pattern = "^set2rbind"))

trg_all_ort <- bind_rows(object_names) |> 
  filter(zmays_eg_homolog_ensembl_gene != "") |>
    distinct() |>
      arrange(up_at)

# Add maize ensembl-uniprot info
## read mapping
zm_nam2up = read.table("../../../input-data/go-uniprot/NAM-UP-anno", header=FALSE, stringsAsFactors = FALSE, sep="\t", quote = "")

# # column arrangement
#desired_order <- c('id', 'ensembl', 'desc', 'ens_at', 'ort_conf', 'perc1', 'perc2', 'up_at', 'descr_at')
desired_order <- c('id', 'ensembl', 'desc', 'ens_zm', 'ort_conf', 'perc1', 'perc2', 'up_zm', 'desc_zm')

# join
library(dplyr)
trg_all_ort_zm = inner_join(trg_all_ort, zm_nam2up, by = c("zmays_eg_homolog_ensembl_gene" = "V1"), relationship="many-to-many") |>
  rename(ensembl = ensembl_gene_id,
         id = up_at,
         ens_zm = zmays_eg_homolog_ensembl_gene,
         perc1 = zmays_eg_homolog_perc_id,
         perc2 = zmays_eg_homolog_perc_id_r1,
         ort_conf = zmays_eg_homolog_orthology_confidence,
         desc = description,
         up_zm = V2,
         desc_zm = V3) |>
    select(all_of(desired_order))
```

```{r source node info}
# read ortholog info
ortho=read.table('../../../input-data/at-ortho_rhythmic-process/mart-all-gramene_up-zea', sep='\t', header=FALSE, stringsAsFactors = FALSE)

#colnames(ortho)=c("ensembl", "id", "ens_at", "ort_conf", "perc1", "perc2", "up_at", "descr_at")
colnames(ortho)=c("ens_zm", "up_zm", "ensembl", "ort_conf", "perc1", "perc2", "id", "desc")

# read maize annotation
anno = read.table("../../../input-data/go-uniprot/NAM_descr_best", header=FALSE, stringsAsFactors = FALSE, sep="\t", quote = "")

# colnames(anno) <- c("ensembl", "desc")
colnames(anno) <- c("ens_zm", "desc_zm")

src_all_ort_zm = inner_join(ortho, anno, by = join_by(ens_zm)) |>
  select(all_of(desired_order)) # order defined in `target node info` chunk

```

```{r join nodes}

x <- unique(rbind(src_all_ort_zm, trg_all_ort_zm))

# read cluster info
clust = read.table("../results/network/c13_m1.35_sort", header=TRUE, stringsAsFactors = FALSE, sep='\t', quote = "")

colnames(clust)=c("ens_zm", "cluster", "member")

# join using maize ensembl ID

library(dplyr)
node.info <- left_join(x, clust, by = join_by(ens_zm)) |>
  relocate(cluster, member, .after = desc)

# TRZEBA WYKLUCZYĆ TE SRC, KTÓRE SĄ ZDUPLIKOWANE W RÓŻNYCH KLASTRACH
# exclude At proteins present in different clusters
## remove NAs
noNA = node.info |> 
  select(id, cluster) |> 
    filter(!is.na(cluster))

## remove duplicates
nn_unique <- noNA |> distinct(id, cluster, .keep_all = TRUE) # takes into account both columns

## keep only IDs in only one cluster
nodes2save <- nn_unique |> 
  group_by(id) |> 
    filter(n_distinct(cluster) == 1)

# Select matching rows from node.info
node.info.filtered = inner_join(nodes2save, node.info, by = join_by("id", "cluster"), relationship = "one-to-many")

## UPs which have no cluster and do not have cluster in other row (i.e. nodes have only NA in cluster)

node.naonly = node.info |> 
  select(id, cluster) |>
    distinct(id, cluster, .keep_all = TRUE) |>
      group_by(id) |>
        filter(n() == 1, is.na(cluster))

# Select matching rows from node.info
node.info.naonly.filt = inner_join(node.naonly, node.info, by = join_by("id", "cluster"), relationship = "one-to-many")

# concatenate
node.info2 = bind_rows(node.info.filtered, node.info.naonly.filt) |>
  ungroup()

# few random rows for manual checking
node.info2 |> sample_n(3)

```

```{r join node edge}
# source node without NAs
src_node.info = node.info2 |> filter(!is.na(cluster))

# join with edge.info
edge.info2 = inner_join(edge.info, src_node.info, by = join_by(up_at1 == id), relationship = "many-to-many")
```

## Split by cluster

```{bash split no filter}
cd ../results/network-rhythm-at-string

# WITHOUT score filtering
awk -v FS='\t' -v OFS='\t' '{print > "cluster_"$18"_ortho_at.txt" }' c13_net_allcl_ortho_at.txt

# save file for first row as header
mv cluster_cluster1_ortho_at.txt header

# add header

for i in 4 6 9 10 13; do
  cat header cluster_${i}_ortho_at.txt > temp && mv temp cluster_${i}_ortho_at.txt ;
done

```

```{bash split}
cd ../results/network-rhythm-at/

# using 400 score cutoff
awk -v FS='\t' -v OFS='\t' '$24>=400 { print > "cluster_"$4"_400_ortho_at.txt" }' c13_net_allcl_ortho_at.txt

# save file for first row as header
mv cluster_cluster1_400_ortho_at.txt header

# add header

for i in 4 6 9 13; do
  cat header cluster_${i}_400_ortho_at.txt > temp && mv temp cluster_${i}_400_ortho_at.txt ;
done

```

:::{.callout-important}
This method gives few connections in networks. The bottleneck is the low number of maize proteins in STRING networks.
However without filtering the results are better - worth checking in Cytoscape.
:::

## Cytoscape networks

```{r}
#| output: false

# load package
library(RCy3)

# check connection to Cytoscape
cytoscapePing()

```

```{r test all-net}
# NETWORK itself

library(dplyr)
node.info = as.data.frame(node.info2 |> arrange(id))

# BYĆ MOŻE NIE PODOBA MU SIĘ, ŻE edge.info i node.info MAJĄ RÓŻNE GENY LUB TO ŻE SĄ DUPLIKATY.
# POTENCJALNE ROZWIĄZNIE JOIN edge.info WG UP SOURCE I TRG Z node.info
# BYĆ MOZE PROBLEM ZE ZNAKAMI Z OPISOW descr

createNetworkFromDataFrames(
  title="Full network - leaf",
  collection="leaf, rhythm, At orthologues",
  nodes=node.info,
  source.id.list="up_at1",
  target.id.list="up_at2",
  edges=edge.info
)
```

```{r nets}
# Clean working directory
rm(list=ls())

# cluster 10 has too few nodes (<4) what prevents Cytoscape analyssi
for (i in c(4, 6, 9, 13)) {
  input_file=paste0("../results/network-rhythm-at/cluster_", i, "_ortho_at.txt")
  
  cl = read.table(input_file, header=TRUE, stringsAsFactors = FALSE, sep='\t', quote = "")
  
# NODE info, rbind columns from source node and target node
src=unique(cl[c('uniprot1', 'ensembl1', 'desc1', 'cluster1', 'member1', "ens_at1", "ort_conf1", "perc1_1", "perc2_1", "up_at1", "descr_at1")])
trg=unique(cl[c('uniprot2', 'ensembl2', 'desc2', 'cluster2', 'member2', "ens_at2", "ort_conf2", "perc1_2", "perc2_2", "up_at2", "descr_at2")])
# names must be the same
names(src) <- c('id', 'ensembl', 'desc', 'cluster', 'member', 'ens_at', 'ort_conf', 'perc1', 'perc2', 'up_at', 'descr_at')
names(trg) <- c('id', 'ensembl', 'desc', 'cluster', 'member', 'ens_at', 'ort_conf', 'perc1', 'perc2', 'up_at', 'descr_at')
# rbind
node.info <- unique(rbind(src, trg))

# sort node info table - in case of UPs multimapping to NAMs select NAM with cluster (sorts numerically, NAs are before numbers. During network generation the last row for each id is selected)
## POTENTIAL PROBLEM, if UP have NAMs from >1 cluster it selects the one with the highest cluster number!!!
## REQUIRE MANUAL CHECKING
ord <- order(node.info[[1]],                 # kolumna 1 – sortowanie leksykograficzne
             as.numeric(node.info[[4]]),    # kolumna 4 – sortowanie numeryczne
             na.last = FALSE)

node.info_srt <- node.info[ord, ]

# EDGE table
edge.info=unique(cl[,!names(cl) %in% c('ensembl1', 'desc1', 'cluster1', 'member1', "ens_at1", "ort_conf1", "perc1_1", "perc2_1", "up_at1", "descr_at1", 'ensembl2', 'desc2', 'cluster2', 'member2', "ens_at2", "ort_conf2", "perc1_2", "perc2_2", "up_at2", "descr_at2")])

# NETWORK itself

createNetworkFromDataFrames(
  title=gsub("../results/network-rhythm-at/", "", input_file),
  collection="leaf, rhythm, At orthologues",
  nodes=node.info_srt,
  source.id.list="uniprot1",
  target.id.list="uniprot2",
  edges=edge.info
)

library(paletteer)
# Użycie palety kolorów
colors <- paletteer_dynamic("cartography::orange.pal", 13)

# Usunięcie ostatnich dwóch znaków - inaczej jest za dużo i kody są niepoprawne (przynajmniej dla tej funkcji)
colors <- substr(colors, 1, 7)

# Ustawienie mapowania kolorów
setNodeColorMapping(
  "cluster",
  c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13"),
  colors,
  mapping.type = "c"
)

setNodeColorBypass(
  t(subset(node.info_srt, cluster == i, select=c("id"))),
  "#78c679"
)

# run "Analyze network" to get degree values
analyzeNetwork()

setNodeSizeMapping(
  "Degree",
  c(1, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100),
  c(20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70),
  mapping.type="c"
)
}
```

## References