---
title: "Prepare network files for analysis in Cytoscape. Genes orthologous to Arabidopsis rhythmic genes retrieved at Orthodb"
format: html
date: "2026-02-19"
bibliography: ../../../refs-stat-anal.bib
execute:
  fig-format: png
  fig.dpi: 300
---

The aim of this code is to filter network files with orthologues of *Arabidopsis thaliana* genes annotated with "Rhythmic process" (GO:0048511) GO term. Orthologues were retrieved from files downloaded from [Orthodb](https://data.orthodb.org/v12/download/odb_data_dump/) and processed in [szkic-orthodb.sh](szkic-orthodb.sh). The script uses clustering results and String network data [@szklarczykSTRINGDatabase20232023] for Cytoscape analysis and visualization [@clineIntegrationBiologicalNetworks2007]. Files are prepared separately for each cluster using 0.7 membership cutoff.
This script uses GO-annotation and associated files from UniProt, prepared in [00-test-filter.qmd](../../../input-data/go-uniprot/00-test-filter.qmd) and [00-GO-and-mapping-UniProt.qmd](../../../input-data/go-uniprot/00-GO-and-mapping-UniProt.qmd) and intermediate files from [10-NAMzUP-leaf-networks-cytoscape-loop.qmd](10-NAMzUP-leaf-networks-cytoscape-loop.qmd).
Here option 1) using NAM - GO is realized

Built with R version `r getRversion()`

## R Preliminaries

Environment cleaning and set-up

```{r packages}
#| eval: false
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Mfuzz")
BiocManager::install("marray")
BiocManager::install("limma")
BiocManager::install("RCy3")
```

```{r}
#| output: false

# Clean working directory
rm(list=ls())
```

## Load data

`c13_m1.35` contains data for clusters.


```{r load}
load("../rdata-saved/mfuzz_c13_m1.35_leafALLsig.RDa")

# gene description
anno = read.table("../../../input-data/go-uniprot/NAM_descr_best", header=FALSE, stringsAsFactors = FALSE, sep="\t", quote = "")
colnames(anno) <- c("gene", "description")

```

## Prepare and export cluster data

```{bash net-dir}
rm -r ../results/network-rhythm-odb/
mkdir ../results/network-rhythm-odb/
```

## Add orthology data to network data

```{r add orthology}
# read network file
zz=gzfile('../results/network/c13_net_allcl.gz') 
net=read.csv(zz, header=FALSE, sep='\t')

colnames(net)=c("uniprot1", "ensembl1", "desc1", "cluster1", "member1", "uniprot2", "ensembl2", "desc2", "cluster2", "member2", "neighborhood", "neighborhood_transferred", "fusion", "cooccurence", "homology", "coexpression", "coexpression_transferred", "experiments", "experiments_transferred", "database", "database_transferred", "textmining", "textmining_transferred", "combined_score")

# read orthologs
ortho=read.table('../../../input-data/at-ortho_rhythmic-process/all4files_joined_othoUP_descr', sep='\t', header=FALSE)

library(dplyr)

#######################
# join using uniprot1

ens1 <- inner_join(net, ortho, by = c("uniprot1" = "V2"), relationship="many-to-many") |>
   rename(up_at1 = V1, 
          descr_at1 = V3
          )

# join using uniprot2

ens2 <- inner_join(ens1, ortho, by = c("uniprot2" = "V2"), relationship="many-to-many") |>
   rename(up_at2 = V1, 
          descr_at2 = V3
          )

sum(duplicated(ens2))
# [1] 0
# no duplicates

# few random rows for manual checking
ens2 |> sample_n(6)

# save to file
write.table(ens2, file="../results/network-rhythm-odb/c13_net_allcl_odb.txt", sep="\t", quote=FALSE, row.names=FALSE, na="")
```

## Split by cluster

```{bash split no filter}
cd ../results/network-rhythm-odb/

# WITHOUT score filtering
awk -v FS='\t' -v OFS='\t' '{print > "cluster_"$4"_odb.txt" }' c13_net_allcl_odb.txt

# save file for first row as header
mv cluster_cluster1_odb.txt header

# add header

for i in 4 5 6 8 10 13; do
  cat header cluster_${i}_odb.txt > temp && mv temp cluster_${i}_odb.txt ;
done

```

```{bash split}
cd ../results/network-rhythm-odb/

# using 400 score cutoff
awk -v FS='\t' -v OFS='\t' '$24>=400 { print > "cluster_"$4"_400_odb.txt" }' c13_net_allcl_odb.txt

# save file for first row as header
mv cluster_cluster1_400_odb.txt header

# add header

for i in 4 6 8 10 13; do
  cat header cluster_${i}_400_odb.txt > temp && mv temp cluster_${i}_400_odb.txt ;
done

```

:::{.callout-important}
This method gives few connections in networks. The bottleneck is the low number of maize proteins in STRING networks.
However without filtering the results are better - worth checking in Cytoscape.
:::

## Cytoscape networks

```{r}
#| output: false

# load package
library(RCy3)

# check connection to Cytoscape
cytoscapePing()

```

```{r nets}
# Clean working directory
rm(list=ls())

# For files WITHOUT filtering
# cluster 10 has too few nodes (<4) what prevents Cytoscape analyssi
for (i in c(4, 5, 6, 8, 10, 13)) {
  input_file=paste0("../results/network-rhythm-odb/cluster_", i, "_odb.txt")
  
  cl = read.table(input_file, header=TRUE, stringsAsFactors = FALSE, sep='\t', quote = "")

# NODE info, rbind columns from source node and target node
src=unique(cl[c('uniprot1', 'ensembl1', 'desc1', 'cluster1', 'member1', "up_at1", "descr_at1")])
trg=unique(cl[c('uniprot2', 'ensembl2', 'desc2', 'cluster2', 'member2', "up_at2", "descr_at2")])

# names must be the same
names(src) <- c('id', 'ensembl', 'desc', 'cluster', 'member', 'up_at', 'descr_at')
names(trg) <- c('id', 'ensembl', 'desc', 'cluster', 'member', 'up_at', 'descr_at')

# rbind
node.info <- unique(rbind(src, trg))

# sort node info table - in case of UPs multimapping to NAMs select NAM with cluster (sorts numerically, NAs are before numbers. During network generation the last row for each id is selected)
## The problem: UP have NAMs from >1 cluster is resolved in 10-NAMzUP-leaf-networks-cytoscape-loop.qmd
## - proteins present in different clusters have been removed.

ord <- order(node.info[[1]],                 # kolumna 1 – sortowanie leksykograficzne
             as.numeric(node.info[[4]]),    # kolumna 4 – sortowanie numeryczne
             na.last = FALSE)

node.info_srt <- node.info[ord, ]

# EDGE table
edge.info=unique(cl[,!names(cl) %in% c('ensembl1', 'desc1', 'cluster1', 'member1', "up_at1", "descr_at1", 'ensembl2', 'desc2', 'cluster2', 'member2', "up_at2", "descr_at2")])

# NETWORK itself

createNetworkFromDataFrames(
  title=gsub("../results/network-rhythm-odb/", "", input_file),
  collection="leaf, rhythm, At orthodb",
  nodes=node.info_srt,
  source.id.list="uniprot1",
  target.id.list="uniprot2",
  edges=edge.info
)

library(paletteer)
# Użycie palety kolorów
colors <- paletteer_dynamic("cartography::orange.pal", 13)

# Usunięcie ostatnich dwóch znaków - inaczej jest za dużo i kody są niepoprawne (przynajmniej dla tej funkcji)
colors <- substr(colors, 1, 7)

# Ustawienie mapowania kolorów
setNodeColorMapping(
  "cluster",
  c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13"),
  colors,
  mapping.type = "c"
)

setNodeColorBypass(
  t(subset(node.info_srt, cluster == i, select=c("id"))),
  "#78c679"
)

# run "Analyze network" to get degree values
analyzeNetwork()

setNodeSizeMapping(
  "Degree",
  c(1, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100),
  c(20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70),
  mapping.type="c"
)
}
```

## References